---
title: "Stumbling with RStan"
linktitle: Stumbling with RStan
date: 2019-09-06
type: docs
reading_time: true
toc: true
categories: ["R", "Stan", "tutorial"]
tags: ["R", "Stan", "trouble", "tutorial"]
menu:
  tutorial:
    parent: Stan
    weight: 2
---

So I was listing to one of my favorite podcasts (DataFramed) a few months ago and a particularly interest guest was on talking.
The guest was Michael Betancourt. 


# Reading about the REALLY cool HMC

HMC is the shorthand for Hamiltonian Monte Carlo.

# Trouble Getting Started

The Getting Started page ('https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started') is extremely thorough and gives some excellent pointerr.
The page boils down to using the following R commands to get started

```{r eval=FALSE}

if (file.exists(".RData")) file.remove(".RData")
install.packages("rstan", repos = "https://cloud.r-project.org/", dependencies = TRUE)


library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

schools_dat <- list(J = 8, 
                     y = c(28,  8, -3,  7, -1,  1, 18, 12),
                     sigma = c(15, 10, 16, 11,  9, 11, 10, 18))

fit <- stan(file = '8schools.stan', data = schools_dat)
```

The Stan file is also incredibly easy to set up by just going up to 'Add File -> Stan File'. 
Super simple...

Or NOT...

I kept getting the following error through back at me.

`Error in compileCode(f, code, language = language, verbose = verbose) : 
  Compilation ERROR, function(s)/method(s) not created! sh: g++: command not found
make: *** [C:/PROGRA~1/R/R-3.6.0/etc/x64/Makeconf:215: file197870bf1a48.o] Error 127
In addition: Warning message:
In system(cmd, intern = !verbose) :
  running command 'C:/PROGRA~1/R/R-3.6.0/bin/x64/R CMD SHLIB file197870bf1a48.cpp 2> file197870bf1a48.cpp.err.txt' had status 1
Error in sink(type = "output") : invalid connection`

Super frustrating...

Anyways, it turns out that after too much time spent Googling and reading stackexchange and GitHub issues that I found how to I was able to fix this annoying startup issue.
There isa file that is created when Stan is being installed that Windows somehow uses to run C++ code from R (I have no clue how this happens...) and there is just a simple setting that I needed to add in order to get Stan to work.
Here is the link to where I found the solution (https://github.com/stan-dev/rstan/issues/633)

Here is how I fixed the issue through R:

1. Uninstall rstan so that I can reinstall it properly
2. Automatically edit the .Makrvars file
3. In this file, make sure the following line is the only line contained in the file: 'CXX14=$(BINPREF)g++ -O2 -march=native -mtune=nativee'
4. Install rstan
5. Start-Up Stan
6. Run Example (takes a minute)

```{r eval=FALSE}

# Step 1
remove.packages('rstan')

# Steps 2-3
dotR <- file.path(Sys.getenv("HOME"), ".R")
if (!file.exists(dotR)) dir.create(dotR)
M <- file.path(dotR, ifelse(.Platform$OS.type == "windows", "Makevars.win", "Makevars"))
if (!file.exists(M)) file.create(M)
cat("CXX14=$(BINPREF)g++ -O2 -march=native -mtune=nativee",
    file = M, sep = "\n", append = TRUE)

# Step 4
if (file.exists(".RData")) file.remove(".RData")
install.packages("rstan", repos = "https://cloud.r-project.org/", dependencies = TRUE)

# Step 5
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)


# Step 6
schools_dat <- list(J = 8, 
                    y = c(28,  8, -3,  7, -1,  1, 18, 12),
                    sigma = c(15, 10, 16, 11,  9, 11, 10, 18))

fit <- stan(file = '8schools.stan', data = schools_dat)

```

Note: I need to come back and edit this content...