---
title: "Introduction to R and Simulating Data"
linktitle: Intro to R and Simulation
date: 2019-01-15T21:13:14-05:00
type: docs
toc: true
#reading_time: 
categories: ["R", "simulation", "tutorial"]
tags: ["R", "simulation", "tutorial"]

menu:
  tutorial:
    parent: General
    weight: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Before we get started with an introduction to simulation, we need to make sure that R and R Studio are downloaded on your computers. 
R is an open source and free statistical programming language. 
R utilizes user built 'packages.'
The package library is huge and you can find a package that does nearly anything you need it to and a large number of these are regularly updated. 
We will talk about packages a lot over the never few years so this is kind of tertiary.

Anyways, to install R follow these steps:

1. Go to https://www.r-project.org/
2. Click on 'to download R'
3. Choose a 'CRAN mirror', a CRAN mirror is where your computer knows to look on the internet to download the software. I suggest https://cran.revolutionanalytics.com/ which is based out of Dallas. 
4. Select the download that is approriate for your computer.
5. Then scroll down to the download: R-3.5.1.pkg (mac) or install for the first time (windows).

These steps will help you install R, next we need to download R Studio. 
R Studio is an interface to use R that is much more friendly and easier to work with. 
You will see this when you open both.

1. Go to https://www.rstudio.com/products/rstudio/download/
2. Scroll down to 'Installers for Supported Platforms' 
3. Select either the windows or mac version

Now that we have R and R studio downloaed, let's dive into an Intro to Simulation.

## Intro to the World of R

R is a statistical programming language.
Because R is a programming language there is a feel of technicality involved with using the software.
One of the major difficulties I have found with helping others get started with R is getting over the initial struggle.
The initial struggle is *OKAY*.
I use R nearly everyday and I am constantly googling how to solve errors and complete my tasks. 
Basically, I want to convey that it's perfectly normal to constantly get error messages and then ask the world of google how to fix your error. 

So, for starters, R can be broken down to being a fancy calculator.
For example, below is a simple case for using R as simple calculator.
We want to find the sum of "1 + 1". 
R will automatically return the result of 2.

```{r}
1 + 1
```

Now, this is a little boring.
One of the major advantages of using a programming language is being able to store different operations into what is called an object.
An object within R is simply a letter or word that represent another symbol or value.
For example, we can store the summation of "1 + 1" into a single letter "x". 
R does this assignment of values to words or letters through two different mechanisms.
You can use "=" or "<-" to make R assign values to objects/letters. 
However, within R it is customary to only use "<-" for basic assignment while the "=" is reserved for arguments within functions (I will get to this in a moment).

```{r}
## Assign the sum of 1 + 1 to x
x <- 1 + 1
## Simply place "x" to see the value stored
x 
```

### Intro to Simple Functions

R is a functional programming language, which means that R is designed with tools that take in values to return a result.
Now, this is super vague and not explicit enough to make sense. 
One of the simplest functions in R is the "print()" function.
The print function takes in a single argument "x". 
This means that based on the value supplied to the function print with the argument x, R will display the actual value(s) stored in x. 
For example, suppose we store the phrase "Hello People" in the object "y".
We can use the print() function to see what the value of "y" is when supplied to the functin.

```{r}
## Store "Hello People" in y
y <- "Hello People"
## Displaythe value of y
print(x=y) ## x is theargument of print, and we want to print y
```
 
This concept of storing values into objects to use in functions is **extremely** important. 
We can utilize this idea in any number of ways we can imagine.
The flexibility to choose how we use the simple rules of R provides us with a basis for manipulating data and doing more complex statistical analyses.
But, for now I will talk a little bit about how R can be used to generate values randomly.

## *Brief* Introduction to Probability

Yes, we are starting with probability. 
Before we can talk about simulating random data, we need to talk briefly about probability.

Let's say we have a quarter. 
A quarter has two sides, HEADS and TAILS.
These two outcomes are possibly if we flip a coin once. 
A coin is called 'fair' if both sides have the same likelihood of occuring. 
This means that over repeated flips of our coin, we would expect that each outcome (H vs. T) to occur approximately the same number of times. 
Therefore, if a coin is fair, each outcome H and T will occur 50\% of the time.
In other words, each outcome has a 0.5 probability of occuring.

This is definition note: a probability is strictly between 0 and 1. If a probability if said to be outside this range, then the value cannot be a probability. 

Probability is based on taking a *sample* from a *population*. 
A population is what is often of interest, but we rarely have access to the entire population. 
The US census is an attempt to get information from the entire US population. 
Because of cost and time, we take sufficiently large samples from a population to gain insight into the population.
For example, to empirically test whether a coin is fair, we need to flip a coin many times generate the number of the times each outcomes occurs. 
If we flip a coin 1000 times and 500 of these turned out to fair, we likely have good evidence that our coin is fair. 
In contrast, if we flipped a coin 3 times and all 3 turned out to heads then the question becomes whether or not we ahve enough evidence to conclude that the coin is *not* fair.

## Simulating a Datum

In R, we can generate a single sample from a population very easily. 
Take the example of a coin flip.
We can specify a population with two outcomes Heads and Tails with each outcome begin equally likely.
From this population, we then take a single draw to obtain a datum.

```{r datum-sim}

# Specify the possible outcomes. We do this by telling R to make an object with two values inside. This is done by using the command  c(.). Values within the () are stuck together
population <- c('HEADS', 'TAILS')

# Specify the Probability of each outcome in the population. Each value in the c(.) must align with the corresponding outcome in the population object. 
outcomeProb <- c(.5, .5)

# To generate the random draw, we use the following:
# x = defined population
# size = number of sample 
# replace = TRUE/FALSE? Should sampling be done with replacement?
sample(x=population, size=1, prob = outcomeProb, replace = T)

```

## Simulating Data

Now, lets simulate more than one datum. We will now simulate 10 coin flips:

```{r}
# To do so, we simply adjust the size to 10 instead of 1
sample(x=population, size=10, prob = outcomeProb, replace = T)

```

